---
title: "Getting Started with deriva"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with deriva}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 70)
```

```{r load}
library(deriva)
```

## What is concept drift?

A machine learning model trained on historical data operates under the implicit assumption that the data-generating process remains stable over time. When this assumption breaks down — because user behaviour shifts, sensor calibration drifts, or the world simply changes — the model's predictions degrade without any obvious error being raised. This phenomenon is called **concept drift**.

Monitoring for drift requires a *drift detector*: an algorithm that reads a stream of per-observation signals (typically prediction errors) and raises a flag when the signal's distribution has changed significantly. The `deriva` package provides a tidy interface to a catalogue of 22 such detectors, designed to compose naturally with the tidymodels ecosystem.

## Quick start

The one-shot shortcut `detect_drift()` runs a detector over an existing column and returns the data annotated with `.warning` and `.drift` flags.

```{r quickstart}
# Simulate a stream: 500 stable observations, then 500 with higher error rate
stream <- sim_drift_stream(n_pre = 500, n_post = 500,
                           p_pre = 0.05, p_post = 0.30,
                           seed = 42)

result <- detect_drift(stream, .col = error, method = "ddm")

# Where was drift flagged?
subset(result, .drift)
```

The detector correctly identifies the distributional change after the known drift point (observation 500).

## The deriva interface

`deriva` follows the same three-verb pattern as tidymodels: **specify → fit → advance**.

### 1. Specify a detector

`drift_detector()` creates an inert specification — no computation happens here.

```{r spec}
spec <- drift_detector("ddm", min_instances = 30)
spec
```

Pass method hyperparameters as named arguments. Unknown parameters raise an informative error.

### 2. Fit on a baseline

`fit()` runs the detector over the **baseline period** — the stable window against which future observations are compared.

```{r fit}
baseline <- sim_drift_stream(n_pre = 300, n_post = 0, seed = 1)

fitted <- fit(spec, baseline, signal = error)
fitted
```

The fitted object is **immutable**: it stores the internal engine state after processing the baseline, ready to receive new data.

### 3. Advance over new batches

`advance()` feeds a new batch to the detector and returns a **new** fitted object with the state updated and the annotated batch appended to the history. The original object is not modified.

```{r advance}
batch1 <- sim_drift_stream(n_pre = 200, n_post = 0,   seed = 2)
batch2 <- sim_drift_stream(n_pre = 0,   n_post = 300, p_post = 0.35, seed = 3)

fitted2 <- advance(fitted,  batch1)
fitted3 <- advance(fitted2, batch2)
fitted3
```

Batches can be any size — including a single observation for true streaming use.

### 4. Inspect results

**`augment()`** returns the full annotated history as a tibble.

```{r augment}
history <- augment(fitted3)
tail(history[, c("t", "error", ".phase", ".warning", ".drift")], 10)
```

**`tidy()`** extracts the detected drift points.

```{r tidy}
tidy(fitted3)
```

**`glance()`** gives a one-row summary.

```{r glance}
glance(fitted3)
```

**`autoplot()`** plots the running mean of the signal with warning (orange) and drift (red) markers, and a dotted line separating baseline from stream.

```{r autoplot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6.5, fig.height = 3.5}
library(ggplot2)
autoplot(fitted3)
```

## Bridging from tidymodels

In a real workflow, the signal column comes from model predictions, not a simulation. `add_prediction_error()` converts the output of tidymodels' `augment()` (which contains truth and estimate columns) into a `.error` column that drift detectors can consume.

```{r bridge}
# Simulate tidymodels augment() output for a classifier
predictions <- data.frame(
  time     = 1:8,
  truth    = factor(c("yes","no","yes","yes","no","yes","no","yes")),
  .pred_class = factor(c("yes","no","yes","no" ,"no","no" ,"no","yes"))
)

add_prediction_error(predictions, truth = truth)
```

For regression problems, `.error` is the absolute prediction error; for classification it is a 0/1 mismatch indicator.

## Distribution-based detectors

Some detectors monitor the distribution of a numeric stream directly, without requiring labelled errors. These `signal_type = "distribution"` methods (such as `"kswin"` and `"adwin"`) expect a continuous input column.

```{r kswin}
cont_stream <- sim_dist_stream(
  n_pre = 500, n_post = 500,
  mean_pre = 0, mean_post = 2,
  seed = 99
)

detect_drift(cont_stream, .col = value, method = "kswin") |>
  subset(.drift) |>
  head()
```

## Available methods

`deriva` ships with 22 drift detectors across two signal types.

| Signal type | Methods |
|---|---|
| `"error"` (0/1 or continuous error) | `ddm`, `eddm`, `hddm_a`, `hddm_w`, `ewma`, `rddm`, `stepd`, `fhddm`, `fhddms`, `mddm_a`, `mddm_e`, `mddm_g`, `wstd`, `ftdd`, `fpdd`, `fsdd`, `cusum` |
| `"distribution"` (numeric stream) | `kswin`, `adwin`, `page_hinkley`, `seed`, `seqdrift2` |

Use `drift_detector("<method>")` to inspect default hyperparameters for any method.

## Summary

The core `deriva` workflow is:

```r
drift_detector("ddm") |>          # specify
  fit(baseline, signal = error) |> # learn reference level
  advance(new_batch)               # update state, persist flags
```

Supplementary verbs — `augment()`, `tidy()`, `glance()`, `autoplot()` — follow the tidymodels convention and make it straightforward to inspect, summarise, and plot detection results at any point in the stream.
