---
title: "Distribution-Based Drift Detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distribution-Based Drift Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load}
library(deriva)
```

## Error-based vs. distribution-based

`vignette("deriva")` covers `signal_type = "error"` methods (DDM and friends): they need a labelled 0/1 (or continuous) error signal, usually built from a model's predictions with `add_prediction_error()`.

`signal_type = "distribution"` methods are different: they watch a raw numeric stream directly — no labels, no baseline error rate — and flag a change in the stream's distribution itself. This is the right family when you want to monitor an input feature or a sensor reading for drift, not just a model's error.

## The running example: a drifting sensor

`sensor_monitoring`, shipped with the package, is a numeric sensor-reading stream: 500 stable observations centred at 0, then 500 after the sensor drifted out of calibration and the mean shifted to 2.

```{r data}
head(sensor_monitoring)
```

## Detecting the shift with KSWIN

`"kswin"` (Kolmogorov-Smirnov Windowing) is a `signal_type = "distribution"` method: it repeatedly compares a recent window of the stream against an older one with a Kolmogorov-Smirnov test.

KSWIN is stochastic — it draws a random sub-sample from its window at every test. Pass `seed` to `drift_detector()` (or, as here, straight into `detect_drift()`, which forwards it) to get a result that is reproducible and does not depend on the state of your session's random number generator.

```{r kswin}
result <- detect_drift(sensor_monitoring, .col = value, method = "kswin", seed = 7)

subset(result, .drift)
```

No detection fires in the 500 stable observations. Unlike DDM, which settles into a new stable state after one detection, KSWIN keeps comparing windows as they slide past the change point — so it fires **several times** while its window catches up to the new distribution, not just once. That is expected behaviour for a windowed method, not noise: every one of these detections comes after the true drift point, as the window repeatedly re-compares against the now-shifted data.

## Other distribution-based methods

`"adwin"` (Adaptive Windowing) solves the same problem with an adaptively-sized window instead of a fixed one, and is deterministic — no `seed` needed:

```{r adwin}
detect_drift(sensor_monitoring, .col = value, method = "adwin") |>
  subset(.drift)
```

On this stream ADWIN settles the way DDM did in `vignette("deriva")`: a single detection, no repeated firing while the window catches up. Which method fires once versus several times is a property of the algorithm, not of one being more "correct" than the other — see `drift_detector("adwin")` for its hyperparameters, and `vignette("deriva")` for the full tidy workflow (`fit()`/`advance()`/`augment()`/...), which works identically for distribution-based methods.
